### 2020-06-10 chapters/sota.tex
Point outlier taxonomy, Partial Subsequence outlier taxonomy

\section{Outlier detection in time-series data}
Blázquez-García et. al review state-of-the-art outlier detection techniques in the context of time-series data and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}. From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey} \cite{Aguinis-Outlier-survey} \cite{Chandola-Outlier-survey-2009} \cite{Hodge-Outlier-survey} \cite{Xu-outlier-survey-2019} we chose to adopt the taxonomy presented in \textit{"A review on outlier/anomaly detection in time-series data"} by Blázquez-García et. al due to its focus on time-series data.

The authors define a time-series as \textit{"observations that have been recorded in an orderly fashion and which are correlated in time"}. Regarding the concept of \textit{outlier} the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}. This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is \textit{“an observation which deviates so much from other observations as to arouse suspicions that it was
generated by a different mechanism"} \cite{Hawkins-Outliers}. Furthermore, Blázquez-García et. al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers. Hence, outlier detection refers to the process of finding anomalies in data. An anomaly relates to a point in time where the behavior of the system does not match the expected one. Note that under this definition the presence of outliers or anomalies does not necessarily imply a problem.

@@

The \textit{input data} axis refers to the type of time-series the method uses as input. According to this criterion, an outlier detection method is said to work with a univariate time-series or a multivariate one. A univariate time-series refers to a time-series consisting of a list of values generated by a single time-dependent variable throughout time. To quote the authors, a univariate time-series is \textit{"an ordered set of real-valued observations, where each observation is recorded at a specific time"}. On the other hand, a multivariate time-series has more than one time-dependent variable. Most importantly, not only is each variable time-dependant but also may have dependencies on other variables. With that in mind, the authors define a multivariate time-series as \textit{"an ordered set of k-dimensional vectors, each of which is recorded at a specific time"}. Each \textit{k-dimensional} vector consists of \textit{k}  time-dependent variables, possibly correlated.

@@

The third and final axis is the \textit{outlier type} the method aims to report. The authors identify three distinct outlier types: point outliers, subsequence outliers and time-series outliers. A point outlier corresponds to a single data point in time that deviates from the expected value. These point outliers can be uni or multivariate, depending on the nature of the method. A univariate point outlier is a single time-dependent variable that stands out from the rest of the time-series. On the other hand, a multivariate point outlier takes into account multiple time-dependent variables in comparison to the remaining time-series. Subsequence outliers refer to a sequence of points in time that together show anomalous behavior, but may not be individually considered outliers. Similarly to point outliers, a subsequence outlier may take into account a single time-dependent variable or multiple ones. The authors classify them as univariate subsequence outliers and multivariate subsequence outliers, respectively. Lastly, time-series outliers refer to entire time-series being outliers themselves. This type of outliers can only be detected if the input data is multivariate.

Kifer et al. 2 window paradigm \\
ref window -> some data\\
sliding window -> current data\\
apply distance metric (like jsd) to both windows if above threshold alert and refill reference (adapts to new distro)\\
input data: uni or multi variate\\
method nature: univariate (doenst check correlation)\\
outlier type: subsequence outlier (window)\\

ExtraN\\
point outlier?\\
density based clusters\\
distance based outliers\\
predictability/forecast property based\\

@@

DeepAnT\\
point outlier, model based, prediction\\
black box DL algorithm\\
explain properties\\
use what authors say in their review\\

>>>>

%Probably say we analyze outlier methods and why aggregations? relate to the approaches thought of to solve the problem and the sections

\section{Outlier detection in time-series data}
Blázquez-García et. al review state-of-the-art outlier detection techniques and present a taxonomy based on the main aspects of outlier detection methods \cite{Blazquez-Garcia-Review-Anomaly-Detection}. From multiple reviews on these kind of methods \cite{Aggarwal-Outlier-survey} \cite{Aguinis-Outlier-survey} \cite{Chandola-Outlier-survey-2009} \cite{Hodge-Outlier-survey} \cite{Xu-outlier-survey-2019} we chose to adopt the taxonomy presented in \textit{"A review on outlier/anomaly detection in time-series data"} by Blázquez-García et. al due to its focus on time-series data.

A time-series is a collection of data points indexed by time. Accordingly, the authors define a time-series as \textit{"observations that have been recorded in an orderly fashion and which are correlated in time"}. Regarding the concept of \textit{outlier}, the authors state that \textit{"outliers can be thought of as observations that do not follow the expected behavior"}. This definition is similar to the concept of outlier provided by Hawkins in 1980 where he states that an outlier is \textit{“an observation which deviates so much from other observations as to arouse suspicions that it was
generated by a different mechanism"} \cite{Hawkins-Outliers}. Furthermore, Blázquez-García et. al allude to the fact that outliers are often referred to in the literature as \textit{"anomalies, discordant observations, discords, exceptions, aberrations, surprises, peculiarities or contaminants"} which for the purpose of their review and this Thesis all refer to outliers. Hence, outlier detection refers to the process of finding anomalies in data. An anomaly relates to a point in time where the behavior of the system does not match the expected one. Note that under this definition the presence of outliers or anomalies does not necessarily imply a problem.

@@

The \textit{input data} axis refers to the type of time-series the method uses as input. According to this criterion, an outlier detection method is said to work with a univariate time-series or a multivariate one. A univariate time-series refers to a time-series consisting of a list of values generated by a single time-dependent variable throughout time. To quote the authors, a univariate time-series is \textit{"an ordered set of real-valued observations, where each observation is recorded at a specific time"}. On the other hand, a multivariate time-series has more than one time-dependent variable. Most importantly, each variable is not only time-dependant but may also have dependencies on other variables. With that in mind, the authors define a multivariate time-series as \textit{"an ordered set of k-dimensional vectors, each of which is recorded at a specific time"}. Each \textit{k-dimensional} vector consists of \textit{k}  time-dependent variables, possibly correlated.

@@

The third and final axis is the \textit{outlier type} the method aims to report. The authors identify three distinct outlier types: point outliers, subsequence outliers and time-series outliers. A point outlier corresponds to a single data point in time that deviates from the expected value. These point outliers can be uni or multivariate, depending on the nature of the method. A univariate point outlier is a single time-dependent variable that stands out from the rest of the time-series. On the other hand, a multivariate point outlier takes into account multiple time-dependent variables in comparison to the remaining time-series. Subsequence outliers refer to a sequence of points in time that together show anomalous behavior, but may not be individually considered outliers. Similarly to point outliers, a subsequence outlier may take into account a single time-dependent variable or multiple ones. The authors classify them as univariate subsequence outliers and multivariate subsequence outliers, respectively. Lastly, time-series outliers refer to entire time-series being outliers themselves. This type of outliers can only be detected if the input data is multivariate, so as to be able to compare multiple time-series.

\subsection*{Univariate point outlier detection}

The authors split point outlier detection methods into uni and multivariate, as seen in Figure \ref{fig:outlier-taxonomy}. Furthermore, they split univariate point outlier detection methods into three broad categories: model-based, density-based and histogramming based \ref{fig:point-outlier-taxonomy}.

\begin{figure}[!htb]
  \begin{center}
    \includegraphics[scale=1]{figures/taxonomy-point-outlier-uni.png}
    \caption{Point outlier detection in univariate time-series taxonomy}
    \label{fig:point-outlier-taxonomy}
  \end{center}
\end{figure}

The univariate point outlier detection methods based on models are the most common. For each timestamp of the time-series, the models compute the expected value for the time-dependent variable and compare it with the observed value. If the difference between the expected value and the observed one is above a certain threshold the data point at the current timestamp is considered an outlier. Mathematically, given the expected value $e\textsubscript{t}$, the actually observed value $x\textsubscript{t}$ and the user-defined threshold of $\alpha$, $x\textsubscript{t}$ is considered an outlier at timestamp \textit{t} if and only if $|$x\textsubscript{t}$-$e\textsubscript{t}$|>\alpha$. Additionally, the authors split the model-based category into estimation and prediction models. The main difference is that estimation model-based methods use past and future data while prediction model-based methods use only past data and infer current and future values. Because of this, prediction model-based outlier detection methods can be used in real-time to process streaming time-series, while estimation based methods can't.

Density-based univariate point outlier detection methods rely on clustering of data points and neighbor counting. These methods consider a datum an outlier if it has less than $\alpha$ neighbors. Distance between two points is usually measured using Euclidean distance. Two points are said to be neighbors if they are close to each other --- \textit{i.e.} the distance between them is less than or equal to a given threshold $\alpha$.

Quoting the authors, histogramming univariate point outlier detection methods are \textit{"based on detecting the points whose removal from the univariate time series results in a histogram representation with lower error than the original"}.
%TODO: explain this better, paper not clear to me

\iffalse
ExtraN\\
point outlier\\
density-based outliers\\
predictability/forecast property based\\

@@

DeepAnT\\
point outlier, model based, prediction\\
black box DL algorithm\\
explain properties\\
use what authors say in their review\\
\fi

\subsection*{Univariate subsequence outlier detection}
Similarly to point outlier detection methods, the authors split subsequence outlier detection methods into uni and multivariate ones \ref{fig:outlier-taxonomy}. Furthermore, the authors split univariate subsequence outlier detection methods into five different axes as seen in Figure \ref{fig:subsequence-outlier-taxonomy}.

\begin{figure}[!htb]
  \begin{center}
    \includegraphics[scale=0.8]{figures/taxonomy-subsequence-outlier-uni.png}
    \caption{Subsequence outlier detection in univariate time-series taxonomy}
    \label{fig:subsequence-outlier-taxonomy}
  \end{center}
\end{figure}

Methods in the discord detection category compare each subsequence in the time-series with the other subsequences within the time-series in order to find out the most unusual ones. Each subsequence is compared to the others using a distance metric, for instances the Euclidean distance. The brute-force way of accomplishing this requires a pairwise comparison of all the subsequences, effectively resulting in a quadratic time complexity. These methods find the most unusual subsequence in a time-series. However, as the authors put it, these type of methods \textit{"do not have a reference of normality or a threshold"} and hence they do not accurately report outliers.

Univariate subsequence outlier detection methods based on dissimilarity are identical to the previous discord detection based methods but use a reference of normality. The representations used for the subsequences and the reference of normality vary among methods in this class of outlier detection methods but they are based on comparison of the reference or normality subsequence with the ones being analyzed for outlier detection. Once again, the metric used to measure the distance between the reference period subsequence and the other ones depends entirely on the method and subsequence aggregation or representation. Finally, when the measured distance is above a certain threshold, these methods classify the subsequence as an outlier or an anomaly.

Prediction model-based univariate subsequence outlier detection methods assume that the reference or normality period is composed of past events. Similarly to univariate point outlier detection methods based on prediction models, these models train on past data and make predictions of the future. Subsequences that do not resemble the prediction made by the model are classified as outliers.

\iffalse
Kifer et al. 2 window paradigm \\
ref window -> some data\\
sliding window -> current data\\
apply distance metric (like jsd) to both windows if above threshold alert and refill reference (adapts to new distro)\\
input data: uni or multi variate\\
method nature: univariate (doenst check correlation)\\
outlier type: subsequence outlier (window)\\
\fi
